The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-08-07 paper.qmd
finished vague question and added risk-aversion paragraph

### Trust and confidence
Trust is a by-product of displaying uncertainty and it commonly measured in uncertainty evaluation studies [@Hullman2019]. Considering trust, and not transparency, as the metric of importance in uncertainty communication can lead to a questionable subtext that argues against transparency, something that has been noticed by several other authors [@Spiegelhalter2017; ONeill2018]. Science communication should be primarily concerned with accuracy.

@@

@@

### Questions that attempts to capture signal-suppression
There is also a swath of studies that are aware a question that boils down to a value extraction experiment, or a question that should be answered by literally ignoring uncertainty information is not what we want when we consider uncertainty visualisations. These papers often try to ask a question that should utilise both the uncertainty and signal in the response, however this is rarely what actually occurs. This method typically results in in cryptic or confusing questions that create a large amount of noise on the interpretation side of the analysis [@Hullman2016].

Some authors opt for asking slightly vague questions that imply a use of uncertainty, but compare it to a ground truth that is very specific. @Ibrekk1987 asked participants for the "best estimate" which was evaluated in accuracy by comparing it to the mean, however the "best estimate" depends on the loss function we are using, and a loss function of minimised error was not implied by the question. @Hofmann2012 showed two distributions in 20 different visualisations (a line-up protocol) using a jittered sample, a density plot, a histogram, and a box plot and asked participants. Participants were asked to report in which of the plots was "the blue group furthest to the right" The experiment set up is shown in fig-right. The participants answers were then compared to a ground truth where the correct plot had a blue distribution with a right shifted mean. By comparing the results to a ground truth statistic and marking participants as "wrong" or "right", the error from the participants that had an alternative interpretation to the concept of "furthest right" was conflated with the error from a the visualisation choice. These papers make it unclear if the participants got the answers wrong because they misunderstood the question or because of something related to the plot. Therefore, this method leads to inconclusive results about the plot design, and is not advised.

Another method used by authors is to ask a deterministic question about a random event. @Padilla2017 provided participants with a visualisation of the cone of uncertainty and asked then to "decide which oil rig will receive more damage based on the depicted forecast of the hurricane path". The cone of uncertainty provides a 60% confidence interval for the location of the eye of a hurricane, which allows us to know the area where the eye of the storm will go, it does not given any information about the intensity of a storm, the size of a storm, or even if a location will be hit. This inclusion of determinism seems to cause the authors to stumble themselves, as they are not consistent with their assumptions. In their first experiment @Padilla2017 indicated the correct answer was to assume that the storm was equally intense no matter how far from the centre of the distribution an oil rig was, however answering their third experiment correctly hinged on assuming the intensity of the storm at a particular point (which in this experiment they phrased as damage) *does* change in intensity as you move away from the centre of the distribution. Given these conflicting assumptions, it is unclear how the participants were supposed to adjust the probabilistic path information to answer a deterministic question about which oil rig would receive the most damage. Other authors have commented on the complexity of communicating hurricane risk because the path, storm surge and wind speed are all important and cannot be ignored [@Spiegelhalter2017]. The flip side of this is asking participants for a deterministic answer to a probabilistic question. @Correll2014 asked participants "how likely is candidate B to win the election?" when the two distributions indicated voter preference. Participants were not able to answer the question about likelihood in term of probability, but were instead given seven options from 1=Outcome will be most in favour of A to 7=Outcome will be most in favour of B. The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorporate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly.

### Identifying heuristics
- Things that are true for both normal visualisation and uncertainty, but of particular interest for uncertainty

@@

>>>>

### Trust, confidence, and risk aversion
Trust is a by-product of displaying uncertainty and it commonly measured in uncertainty evaluation studies [@Hullman2019]. Considering trust, and not transparency, as the metric of importance in uncertainty communication can lead to a questionable subtext that argues against transparency, something that has been noticed by several other authors [@Spiegelhalter2017; ONeill2018]. Science communication should be primarily concerned with accuracy.

@@

Unfortunately, experiments where the best response comes from participants who outright ignore the uncertainty information are not uncommon. A simple example of this is experiments where participants are expectd to extract the mean of the distribution. In these cases, we would expect the best performing visualisation to be one that does not include uncertainty at all, as it would simply show the mean. Similarly, many uncertainty visualisations are evaluated on how much risk aversion the participants display [@Hullman2019]. Additionally, comparing participant responses to that of a ration agent has been suggested as a benchmark for uncertainty visualization experiments [@wu2023rational]. This standard seems to go against the purpose of uncertainty visualisation as a rational agent, *by definition* should *ignore* uncertainty information. Risk aversion is considered to be irrational because it means the economic agent *is* considering the uncertainty information. The only cases where a rational agent should not ignore the uncertainty information is when it is required to optimise their utility function, which would mean the uncertainty is not acting as noise, it is acting as signal.

@@

For these reasons we do not believe trust, confidence, or risk-aversion are useful measures to evaluate uncertainty visualisations. While they are designed to capture the secondary effects of uncertainty, using them as primary measures of visualisation performance incentives the use of visualisations that *do not* include uncertainty.

### Questions that attempts to capture signal-suppression
There is a collection of studies that seem to be aware of the issues behind using trust or value extraction to evaluate uncertainty visualisations. These studies try to measure the effect of some kind of "single integrated value" but the methodology is often ad-hoc with varying levels of success.

The first method is what we call the "vague question" approach. The authors of these studies will ask the participants a question that implies that they should use  a use of uncertainty, but compare the readers answer to a very specific ground truth. This means the participants are being *evaluated* as though they are performing a value extraction task, but they are not being asked the *questions* that are asked in the value extraction task. Ultimately this approach results in strangely cryptic questions that create a large amount of noise due to the varied interpretation of the questions [@Hullman2016]. For example @Hofmann2012 showed study participants 20 plots, where each plot displayed two distributiosn (as a pair of jittered samples, density plots, histograms, or box plots) and asked them to identify the plot where "the blue group furthest to the right". The mean of the two groups was used to decide which group was actually "furthest to the right" and the participants were evaluated against that ground truth. In another example @Ibrekk1987 asked participants for the "best estimate", but the participant's responses were evaluated against the mean of the distribution. Ultimately, the term "best" is up to the users interpretation, and the estimate that minimized the sum of squared errors was not implied by the question. This vague question approach leads to inconclusive results, as we are left unclear if it was the phrasing of the question or the plot design that caused the participants to answer incorrectly.

A variation of the "vague question" problem, is a series of studies that ask questions that are impossible to answer using the information given to the participants. These studies frequently expect participants to give deterministic answers for probabilistic questions. For example @Correll2014 showed participants the distribution of voter preferences for two candidates in an election, and aksed them "how likely is candidate B to win the election?". Participants were not able to answer the question about likelihood in term of probability, but were instead given seven options from 1 = "Outcome will be most in favour of A" to 7 = "Outcome will be most in favour of B". The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorporate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly.  In another example, @Padilla2017 provided participants with a visualisation of the cone of uncertainty and asked then to "decide which oil rig will receive more damage based on the depicted forecast of the hurricane path". The cone of uncertainty provides a 60% confidence interval for the location of the eye of a hurricane, which allows us to know the area where the eye of the storm will go, it does not given any information about the intensity of a storm, the size of a storm, or even if a location will be hit. Other authors have commented on the complexity of communicating hurricane risk because the path, storm surge and wind speed are all important and cannot be ignored [@Spiegelhalter2017]. Interestingly it is unclear how the participants were supposed to answer this question, as @Padilla2017 did three experiments and the first and third experiment had conflicting assumptions in their ground truth. In the first study, participants were required assume to the storm was equally intense reguardless of the probability of the probability of the oil rig being hit (i.e. more likely to be hit *does not* mean more damage) but in the second study required participants to assume the opposite (i.e. more likely to be hit *does* mean more damage). Similar to the vague question problem, these studies seem to be aware that we should be evaluating an uncertainty visualization based on a deterministic observation (such as our example of "does this map have a spatial trend") but are unsure how to incorporate or evaluate it.

While these approaches are certainly a step in the right direction, the experiments end up having far too much noise in their results. Additionally, by having a ground truth, they end up implicitly asking users to either treat uncertainty as a signal, or to ignore it entirely, neither of which is advisable for evaluating graphics on signal-suppression. So, if we cannot ask direct questions about uncertainty, *and* we can't use uncertainty statistics as ground truths in our experiments, how *should* we evaluate uncertainty?

@@

## Identifying heuristics
- Things that are true for both normal visualisation and uncertainty, but of particular interest for uncertainty
